\documentclass{zett}
\input{preambles}
\usetikzlibrary {arrows.meta,backgrounds,fit,positioning}
\title{Graphs}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Degrees}
\label{sec:degrees}

\begin{defn}
  Let $G = (V,E)$ be a simple (no self loops) undirected graph.
  A \emph{neighbor} of a given vertex $v \in V$ is a vertex $u \in V$ such that $\Set{u,v} \in E$.
\end{defn}

\begin{defn}
  Let $G = (V,E)$ be a simple undirected graph.
  The \emph{degree} of a given vertex $v \in V$, denoted as $\deg(v)$, is the number of neighbors of $v$.
\end{defn}

\begin{lem}[Handshake Lemma]
  \label{lem:handshake-lemma}
  Let $G = (V,E)$ be a simple undirected graph.
  \[
    2|E| = \sum_{v \in V}\deg(v)
  \]
\end{lem}
\begin{proof}
  \begin{node}
    Proceed by induction on the number of vertices.
    In the base case, there is no vertex, so there is no edges either.
    In the induction step, there are $k + 1$ vertices.
    By removing an arbitrary vertex $u$ from this graph, we are left with a graph with $k$ vertices.
    By the induction hypothesis,
    \[
      2|E'| = \sum_{v \in V \setminus \Set{u}}\deg(v)
    \]
    where $E'$ is the set of remaining edges after removing $u$.
  \end{node}
  \begin{node}
    Note that $2|E'| + 2\deg(u) = 2|E|$ because we had to remove exactly $\deg(u)$ edges when we removed $u$.
    And $2\deg(u) + \sum_{v \in V \setminus \Set{u}}\deg(v) = \sum_{v \in V}\deg(v)$.
  \end{node}
\end{proof}

\begin{defn}
  A simple undirected graph $G$ is \emph{complete} if every vertex is adjacent to every other vertex.
\end{defn}

\begin{cor}
  The complete graph $K_{n}$ with $n$ vertices has $\frac{n(n-1)}{2}$ edges.
\end{cor}
\begin{proof}
  Since every vertex is adjacent to every other vertex, the degree of each vertex is $n - 1$.
  By \cref{lem:handshake-lemma},
  \[
    2|E| = \sum_{v \in V}\deg(v) = \sum_{v \in V}(n-1) = n(n-1)
  \]
\end{proof}

\section{Graph homomorphisms}
\label{sec:graph-homomorphisms}

\begin{node}
  The word \emph{homomorphism} comes from Greek: \emph{homos} means ``same'' and \emph{morphe} means ``shape.''
  A homomorphism is a structure-preserving mapping.
\end{node}
\begin{node}
  A graph is a set $V$ equipped with a structure (a binary relation $E \subseteq V \times V$).
\end{node}

\begin{defn}
  Let $G = (V,E)$ and $G' = (V',E')$ be graphs.
  A \emph{graph homomorphism} $\varphi : G \to G'$ consists of an underlying function $|\varphi| : V \to V'$ that preserves the relation $E$, i.e., for all $u, v \in V$, if $(u,v) \in E$ then $(|\varphi|(u),|\varphi|(v)) \in E'$.
\end{defn}

\begin{lem}
  Let $G = (V, E)$ be a graph.
  The identity function $\id_{V} : V \to V$ is a graph homomorphism, denoted as $\id_{G} : G \to G$.
\end{lem}
\begin{proof}
  \begin{node}
    Let $u,v \in V$ be any vertices of $G$.
    Suppose that $(u,v) \in E$.
    Then clearly $(\id_{V}(u), \id_{V}(v)) = (u,v) \in E$.
  \end{node}
\end{proof}

\begin{node}
  \label{node:homomorphism-not-just-function}
  A graph homomorphism is not just a function on vertex sets.
  The graph structures are important.
  For example, given a set $V = \Set{A,B,C}$, any subset $E \subseteq V \times V$ is a graph structure on $V$.
  Two obvious graphs are the indiscrete graph $I = (V,\varnothing)$ and the discrete graph $D = (V,V \times V)$.
  The identity function on $V$ is a graph homomorphism $I \to D$, but it is not a graph homomorphism $D \to I$.
\end{node}

\begin{lem}
  Let $G = (V, E)$, $G' = (V', E')$, and $G'' = (V'', E'')$ be graphs and $\varphi : G \to G'$ and $\varphi' : G' \to G''$ be graph homomorphisms.
  The composite $|\varphi'| \circ |\varphi| : V \to V''$ is a graph homomorphism, denoted as $\varphi' \circ \varphi : G \to G''$.
\end{lem}
\begin{proof}
  Let $u,v \in V$ be given.
  Assume that $(u,v) \in E$.
  We need to show that $(|\varphi'| \circ |\varphi|(u),|\varphi'| \circ |\varphi|(v)) \in E''$.
  Since $\varphi$ is a graph homomorphism, $(|\varphi|(u),|\varphi|(v)) \in E'$.
  And since $\varphi'$ is a graph homomorphism, the result follows.
\end{proof}

\section{Graph isomorphisms}
\label{sec:graph-isomorphisms}

\begin{defn}
  Let $G$ and $G'$ be graphs.
  A graph homomorphism $\varphi : G \to G'$ is a \emph{graph isomorphism} if there is a graph homomorphisms $\varphi\inv : G' \to G$ such that $\varphi \circ \varphi\inv = \id_{G'}$ and $\varphi\inv \circ \varphi = \id_{G}$.
\end{defn}

\begin{node}
  \cref{node:homomorphism-not-just-function} has demonstrated that not every bijection is a graph isomorphism because $D$ and $I$ are not isomorphic.
  However, the underlying function of every graph isomorphism is a bijection.
\end{node}

\begin{notn}
  We write $G \iso G'$ when there is a graph isomorphism between the two graphs.
\end{notn}

\section{Graph coloring}
\label{sec:graph-coloring}

\begin{defn}
  Let $G = (V, E)$ be a simple undirected graph, a \emph{$k$-coloring} is a function $\chi : V \to \Set{1,2,\ldots,k}$ such that for any $u,v \in V$, if $\Set{u,v} \in E$ then $\chi(u) \neq \chi(v)$.
\end{defn}

\begin{lem}
  \label{lem:graph-coloring-induced-coloring}
  Let $G = (V,E)$ and $G' = (V',E')$ be graphs.
  A graph homomorphism $\varphi : G \to G'$ together with a $k$-coloring $\chi$ of $G'$ induce a $k$-coloring of $G$.
  \[
    % https://q.uiver.app/#q=WzAsMyxbMCwwLCJWIl0sWzIsMCwiViciXSxbMSwxLCJcXFNldHsxLDIsXFxsZG90cyxrfSJdLFswLDEsInxcXHZhcnBoaXwiXSxbMSwyLCJmIl0sWzAsMiwiZiBcXGNpcmMgfFxcdmFycGhpfCIsMl1d
    \begin{tikzcd}
      V && {V'} \\
      & {\Set{1,2,\ldots,k}}
      \arrow["{|\varphi|}", from=1-1, to=1-3]
      \arrow["\chi", from=1-3, to=2-2]
      \arrow["{\chi \circ |\varphi|}"', from=1-1, to=2-2]
    \end{tikzcd}
  \]
\end{lem}
\begin{proof}
  \begin{node}
    Suppose that $\Set{u,v} \in E$.
    We need to verify that $\chi \circ |\varphi|(u) \neq \chi \circ |\varphi|(v)$.
    Since $\varphi$ is a graph homomorphism, $\Set{|\varphi|(u),|\varphi|(v)} \in E'$.
    Since $\chi$ is a $k$-coloring of $G'$, the result follows.
  \end{node}
\end{proof}

\begin{defn}
  Let $G$ be a simple undirected graph.
  The \emph{chromatic number} $\chi_{G}$ of $G$ is the minimum number such that $G$ has a $\chi_{G}$-coloring.
\end{defn}

\begin{lem}
  Chromatic number is a graph theoretic property, i.e., if $G \iso G'$ and $G$ has chromatic number $k$, then $G'$ also has chromatic number $k$.
\end{lem}
\begin{proof}
  \begin{node}
    By \cref{lem:graph-coloring-induced-coloring}, a $k$-coloring of $G$ and an isomorphism $\varphi : G' \to G$ induces a $k$-coloring of $G'$.
    It suffices to show that $k$ is the minimal number that makes $G'$ colorable.
  \end{node}
  \begin{node}
    Suppose that there is some $k' < k$ such that $G'$ is $k'$ colorable.
    Any $k'$-coloring of $G'$ together with the inverse isomorphism $\varphi\inv : G \to G'$ induce a $k'$-coloring of $G$.
    This contradicts the minimality of $k$.
  \end{node}
\end{proof}

\end{document}